Unifying the communicable disease spreading paradigm with Gompertzian growth

### 2022-07-21 elsarticle/elsarticle.tex
add some to the introduction

\begin{abstract}
Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
We show that the Gompertz curve is incompatible with the traditional communicable disease spreading hypothesis, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor. 
Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
Third, we show that the population acts like a coherent organism under growth/depletion. 
Fourth, we connect the Susceptible-Infected-Recovered (SIR) model with our new theory and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction. We thus connect logistic and Gompertzian growth through a mechanism of higher-order interactions expressed in a closed form.
\end{abstract}

@@

This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves, and are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics. Other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.

What is causing such a discrepancy between these observations and the current theory of communicable diseases \citep{castro2020turning}?

We start by reviewing a basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge. First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I\nonumber,
\end{align}
omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

@@

\section{Conclusion}
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network. In this vein, Richard's parameter, $\theta$, can be related to the number of higher order interactions with the susceptible and the infected in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.

>>>>

\begin{abstract}
Recently, a number of studies have shown that cumulative mortality followed a Gompertz curve in the initial Coronavirus epidemic period, March-April 2020. 
We show that the Gompertz curve is incompatible with the traditional communicable disease spreading hypothesis, and propose a new theory which better explains the nature of the mortality characteristics based on an environmental stressor. 
Second, we show that for the Gompertz curve to emerge, the stressor has to act on everyone simultaneously, rejecting the possibility of a disease propagation stage. 
Third, we show that the population acts like a coherent organism under growth/depletion. 
Fourth, we connect the Susceptible-Infected-Recovered (SIR) model with the Gompertz curve by augmenting the logistic growth equation and show that the SIR model is compatible with Gompertzian growth only when all nodes in the transmission network communicate with infinite speed and interaction. We thus find a novel parametrization bridging logistic and Gompertzian growth.
\end{abstract}

@@

This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves. These observations are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics. What is causing such a discrepancy between these observations and the current theory of communicable diseases \citep{castro2020turning}?

In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}. Many of these animal mortality spikes are not attributable to the coronavirus, but the question still remains why these spikes all occurred within the same time frame. While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory with fewer parameters than the most basic communicable disease model. We will then present a novel bridge between the these two paradigms which connects the logistic growth model with the Gompertzian growth model in an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.

\section{Simple communicable disease models}
We start by reviewing this basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge. First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I\nonumber,
\end{align}
omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

@@

\section{Conclusion}
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network. In this vein, Richard's parameter, $\theta$, related although not identical to the number of higher order interactions with the susceptible and the infected in a SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.
